---
title: "NOAA Storm Database Analysis"
output: html_document
---
#### Events affecting Economy and causing fatalities or injuries

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project analyzes the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to determine the events affecting the most the economy or causing the most fatalities or injuries.

## Data Processing

### Loading required libraries

```{r}
library(dplyr, quietly = TRUE)
library(R.utils, quietly = TRUE)
```

### Downloading the data

```{r}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if (!file.exists("data")) {
  dir.create("data")
}

if (!file.exists("./data/repdata-data-StormData.csv")) {
  if (!file.exists("./data/repdata-data-StormData.csv.bz2")) {
    download.file(fileUrl, destfile = "./data/repdata-data-StormData.csv.bz2")
  }  
  bunzip2("./data/repdata-data-StormData.csv.bz2", "./data/repdata-data-StormData.csv")
}

noaa_full <- read.csv("./data/repdata-data-StormData.csv")
```

### Subsetting and cleaning the data

We keep only the fields pertinent to our analysis:

- EVTYPE: the type of event
- FATALITIES: number of fatalities
- INJURIES: number of injuries
- PROPDMG: damage to properties in USD
- PROPDMGEXP: magnitude for properties damage (“K” for thousands, “M” for millions, “B” for billions)
- CROPDMG: damage to crops in USD
- CROPDMGEXP: magnitude for crops damage (“K” for thousands, “M” for millions, “B” for billions)

```{r}
noaa <- noaa_full %>% select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)
```

As the magnitude for damages amount contains invalid characters and is not practical in this format for the rest of the analysis, we clean the data and convert it to a power of ten we can multiply with the damage amount to get the final damage value in USD. We consider the invalid characters to be missing data having no value.

```{r}
noaa$PROPDMGEXP <- sub("[-?+H123456789]", "0", noaa$PROPDMGEXP, ignore.case = TRUE)
noaa$PROPDMGEXP <- sub("^$", "0", noaa$PROPDMGEXP)
noaa$PROPDMGEXP <- sub("^K$", "1000", noaa$PROPDMGEXP, ignore.case = TRUE)
noaa$PROPDMGEXP <- sub("^M$", "1000000", noaa$PROPDMGEXP, ignore.case = TRUE)
noaa$PROPDMGEXP <- sub("^B$", "1000000000", noaa$PROPDMGEXP, ignore.case = TRUE)
noaa$PROPDMGEXP <- as.numeric(noaa$PROPDMGEXP)


noaa$CROPDMGEXP <- sub("[-?+H123456789]", "0", noaa$CROPDMGEXP, ignore.case = TRUE)
noaa$CROPDMGEXP <- sub("^$", "0", noaa$CROPDMGEXP)
noaa$CROPDMGEXP <- sub("^K$", "1000", noaa$CROPDMGEXP, ignore.case = TRUE)
noaa$CROPDMGEXP <- sub("^M$", "1000000", noaa$CROPDMGEXP, ignore.case = TRUE)
noaa$CROPDMGEXP <- sub("^B$", "1000000000", noaa$CROPDMGEXP, ignore.case = TRUE)
noaa$CROPDMGEXP <- as.numeric(noaa$CROPDMGEXP)
```

```{r}
event_fatalities <- noaa %>%
                    filter(FATALITIES > 0) %>%
                    group_by(EVTYPE) %>%
                    summarise( FATALITIES = sum(FATALITIES, na.rm = TRUE)) %>%
                    arrange(desc(FATALITIES)) %>%
                    head(10)


barplot(event_fatalities$FATALITIES, names.arg=event_fatalities$EVTYPE, 
        ylab="Fatalities", 
        las=3, 
        col = "blue",
        main="Top 10 Events causing most fatalities")

```

```{r}

event_injuries <- noaa %>%
                    filter(INJURIES > 0) %>%
                    group_by(EVTYPE) %>%
                    summarise( INJURIES = sum(INJURIES, na.rm = TRUE)) %>%
                    arrange(desc(INJURIES)) %>%
                    head(10)

barplot(event_injuries$INJURIES, names.arg=event_injuries$EVTYPE, 
        ylab="Injuries", 
        las=3, 
        col = "blue",
        main="Top 10 Events causing most injuries")

```


